Performance of Jellyfish Search Optimiser on Projection Pursuit Problems

### 2024-03-07 optim.qmd
added math notations

[Put the code in, add specifics to this visualisation application]

@@

An intuitive way to measure smoothness would be to find how many continuous derivatives exist. We make use of Sobolev spaces:

[def of sobolev space]

Smoothness would then be the highest p such that the index function belongs to $W^{p,\infty}$. We can make it weaker by considering $W^{p,q}$ Sobolev spaces.

[backup plan when things are not differentiable]

@@

A large squint angle means the function is easy to optimize, because you don’t need to be very very close to the perfect view to see the structure. A small squint angle means that the derivative of the index function can still be very large values near the optimal point.

OR
that the values taken by the index function can be clustered into two clear groups: one small group of high values near the optimum, and a large group of the remaining small values.
OR
the rate of change of gradient? Gradient changes very drastically when it’s near optimal?

[backup plan when things are not differentiable]

@@

Computational complexity (in big O notation, with respect to the sample size) of computing the index function.

>>>>

[Put the pseudo-code in, add specifics to this visualisation application]

@@

An intuitive way to measure smoothness of a function would be to count how many continuous derivatives exist. To help define smoothness in our context, we can make use of the Sobolev spaces: functions $f$ in the Sobolev space $W^{p,\infty}$ have all derivatives of order less than $p$ continuous. Smoothness would then be the highest $p$ such that the index function belongs to $W^{p,\infty}$. We can relax this and consider $W^{p,q}$ Sobolev spaces.

Consider the following definition of partial derivative. Let $U$ be an open subset of $\mathbf{R}^n$ and $f: U \rightarrow \mathbf{R}$. The partial derivative of function $f$ at the point $\mathbf{a} = (a_1, \ldots, a_n) \in U$ with respective to the $i$-th variable $x_i$ is defined as

$$
\frac{\delta}{\delta x_i} f(\mathbf{a}) =  \lim_{h \rightarrow 0} \frac{f(a_1, \ldots, a_{i-1}, a_i +h, a_{i+1}, \ldots, a_n) - f(a_1 , \ldots, a_i, \ldots, a_n)}{h} = \lim_{h\rightarrow 0} \frac{f(\mathbf{a} + h\mathbf{e_i}) - f(\mathbf{a})}{h}.
$$
where $\mathbf{e_i}$ is the unit vector of the $i$-th variable $x_i$.

If the derivative is not well-defined, we propose to approximate the derivative using the following expression:

$$\frac{1}{n_i}\sum_{i} \frac{\|f(\mathbf{a} + h_i\mathbf{e_i}) - f(\mathbf{a})\|^p}{h_i}$$

where $p \in [0,1]$. The choice of $p$ reflects the penalty behaviour. For the application in this paper, we choose $p = 1$.

- $h_i$ is an neighbourhood area around $a$. We can insert different values of $h$ to approximate the above quantity and observe how this quantity changes.

@@

From the literature, it is commonly understood that a large squint angle implies that the function is easy to optimize, because we do not need to be very close to the perfect view to see the structure. A small squint angle means that the derivative of the index function can still be very large values near the optimal point and will rapidly change as we get even closer to the optimum. As such we can observe the second order gradient, which is the rate of change of gradient, over the space we are searching. For some index functions, the second order gradient is not well-defined, we can approximate the second order gradient vector in similar fashion as the above section.

@@

The speed of optimizing an index function can be calculated/measured using the computational complexity (in big O notation, with respect to the sample size) of computing the index function.

### 2024-03-07 optim.tex
added math notations

\begin{keyword}
    projection pursuit \sep optimization \sep 
    jellyfish optimiser
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

{[}Put the code in, add specifics to this visualisation application{]}

@@

An intuitive way to measure smoothness would be to find how many
continuous derivatives exist. We make use of Sobolev spaces:

{[}def of sobolev space{]}

Smoothness would then be the highest p such that the index function
belongs to \(W^{p,\infty}\). We can make it weaker by considering
\(W^{p,q}\) Sobolev spaces.

{[}backup plan when things are not differentiable{]}

@@

A large squint angle means the function is easy to optimize, because you
don't need to be very very close to the perfect view to see the
structure. A small squint angle means that the derivative of the index
function can still be very large values near the optimal point.

OR that the values taken by the index function can be clustered into two
clear groups: one small group of high values near the optimum, and a
large group of the remaining small values. OR the rate of change of
gradient? Gradient changes very drastically when it's near optimal?

{[}backup plan when things are not differentiable{]}

>>>>

\begin{keyword}
    projection pursuit \sep optimization \sep 
    jellyfish optimiser
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

{[}Put the pseudo-code in, add specifics to this visualisation
application{]}

@@

An intuitive way to measure smoothness of a function would be to count
how many continuous derivatives exist. To help define smoothness in our
context, we can make use of the Sobolev spaces: functions \(f\) in the
Sobolev space \(W^{p,\infty}\) have all derivatives of order less than
\(p\) continuous. Smoothness would then be the highest \(p\) such that
the index function belongs to \(W^{p,\infty}\). We can relax this and
consider \(W^{p,q}\) Sobolev spaces.

Consider the following definition of partial derivative. Let \(U\) be an
open subset of \(\mathbf{R}^n\) and \(f: U \rightarrow \mathbf{R}\). The
partial derivative of function \(f\) at the point
\(\mathbf{a} = (a_1, \ldots, a_n) \in U\) with respective to the
\(i\)-th variable \(x_i\) is defined as

\[
\frac{\delta}{\delta x_i} f(\mathbf{a}) =  \lim_{h \rightarrow 0} \frac{f(a_1, \ldots, a_{i-1}, a_i +h, a_{i+1}, \ldots, a_n) - f(a_1 , \ldots, a_i, \ldots, a_n)}{h} = \lim_{h\rightarrow 0} \frac{f(\mathbf{a} + h\mathbf{e_i}) - f(\mathbf{a})}{h}.
\] where \(\mathbf{e_i}\) is the unit vector of the \(i\)-th variable
\(x_i\).

If the derivative is not well-defined, we propose to approximate the
derivative using the following expression:

\[\frac{1}{n_i}\sum_{i} \frac{\|f(\mathbf{a} + h_i\mathbf{e_i}) - f(\mathbf{a})\|^p}{h_i}\]

where \(p \in [0,1]\). The choice of \(p\) reflects the penalty
behaviour. For the application in this paper, we choose \(p = 1\).

\begin{itemize}
\tightlist
\item
  \(h_i\) is an neighbourhood area around \(a\). We can insert different
  values of \(h\) to approximate the above quantity and observe how this
  quantity changes.
\end{itemize}

@@

From the literature, it is commonly understood that a large squint angle
implies that the function is easy to optimize, because we do not need to
be very close to the perfect view to see the structure. A small squint
angle means that the derivative of the index function can still be very
large values near the optimal point and will rapidly change as we get
even closer to the optimum. As such we can observe the second order
gradient, which is the rate of change of gradient, over the space we are
searching. For some index functions, the second order gradient is not
well-defined, we can approximate the second order gradient vector in
similar fashion as the above section.
